---
title: "Predicting Flight Delays and Understanding Contributing Factors"
author: "Untitled"
date: "`r Sys.Date()`"
output: pdf_document
---
### Load Packages
```{r}
library(dplyr)#use for data manipulation
library(car) # use for multicollinearity test (i.e. Variance Inflation Factor(VIF))
library(ggplot2) # use for visualization
library(gridExtra) # To plot multiple ggplot graphs in a grid
library(olsrr) #for multicollinearityuse  test
```

#import dataset
```{r}
Delay_Flight <- read.csv("C:\\Users\\Elite\\Downloads\\archive (1)\\airline_delay.csv")

```

EXporatory data analysis
## check make up of data

```{r}
# Look at the first and last few rows to ensure that the data is read in properly and check for consistency

head(Delay_Flight)
tail(Delay_Flight)


```

```{r}
dim(Delay_Flight)
# There are 3351 observations and 21 variables
```
```{r}
colnames(Delay_Flight)
```


get info about the dataset

```{r}
summary(Delay_Flight)
```
Observations:
- For the variables where the Mean and median  values is very different suggests a skewed distribution and presence of outliers.

```{r}
unique(Delay_Flight)
```

```{r}
unique(Delay_Flight$year)
unique(Delay_Flight$month)
unique(Delay_Flight$carrier)
unique(Delay_Flight$carrier_name)
unique(Delay_Flight$arr_del15)
unique(Delay_Flight$carrier_delay)
unique(Delay_Flight$weather_delay)
unique(Delay_Flight$arr_diverted)
unique(Delay_Flight$arr_cancelled)
unique(Delay_Flight$carrier_delay)


```

```{r}
View(Delay_Flight)

```

```{r}
str(Delay_Flight)
#it can be seen that there would be a need to changed data types for some variables 

```
Descriptive Statistics - the arr_del15  variable is our dependent/target/response variable.

Data Cleaning :

1. Variable conversion : arr_del15 -  * It has a interger type, but character value doesn't mean anything. It's better to change it into a numeric type. 

```{r}
# Convert Interger to numeric
Delay_Flight$arr_del15 <- as.numeric(Delay_Flight$arr_del15)
summary(Delay_Flight$arr_del15)
```
Observations:
- It can be seen now that there appear to be an obvious difference between the median and mean indicating a skewed distribution and possible presence of outliers.
- Also there are missing values which didn't show because it was recorded as interger.
These will be investigated during the visualizations.


2. Coverting data type: arr_flights,arr_cancelled,arr_diverted,arr_delay,carrier_delay,nas_delay,weather_delay,security_delay,late_aircraft_delay
They are Interger variable, but should be numeric type.

```{r}
Delay_Flight$arr_flights <- as.numeric(Delay_Flight$arr_flights)
Delay_Flight$arr_cancelled <- as.numeric(Delay_Flight$arr_cancelled)
Delay_Flight$arr_diverted <- as.numeric(Delay_Flight$arr_diverted)
Delay_Flight$arr_delay <- as.numeric(Delay_Flight$arr_delay)
Delay_Flight$carrier_delay <- as.numeric(Delay_Flight$carrier_delay)
Delay_Flight$weather_delay <- as.numeric(Delay_Flight$weather_delay)
Delay_Flight$nas_delay <- as.numeric(Delay_Flight$nas_delay)
Delay_Flight$security_delay <- as.numeric(Delay_Flight$security_delay)
Delay_Flight$late_aircraft_delay <- as.numeric(Delay_Flight$late_aircraft_delay)
Delay_Flight$year <- as.factor(Delay_Flight$year)


```

Now check structure after converting datatypes

```{r}
str(Delay_Flight)

```

Missing Value Treatment 
```{r}
# Check the missing values in the dataset
sum(is.na(Delay_Flight))

# Check which column has missing values
colSums(is.na(Delay_Flight))
```
As can be seen there are 8 mising values in the arr_flights ,arr_del15,carrier_ct,weather_ct,nas_ct,security_ct,ate_aircraft_ct,arr_cancelled        arr_diverted,arr_delay,carrier_delay, weather_delay,nas_delay,late_aircraft_delay  security_delay variable. We would treat this by replacing all missing values with the median.

```{r}
# missing value imputation
Delay_Flight$arr_flights[which(is.na(Delay_Flight$arr_flights))]<- median(Delay_Flight$arr_flights,na.rm = T)

# check the missing value again. 
sum(is.na(Delay_Flight))
```

```{r}
Delay_Flight$arr_del15[which(is.na(Delay_Flight$arr_del15))]<- median(Delay_Flight$arr_del15,na.rm = T)
sum(is.na(Delay_Flight))







```

```{r}
Delay_Flight$carrier_ct[which(is.na(Delay_Flight$carrier_ct))]<- median(Delay_Flight$carrier_ct,na.rm = T)
sum(is.na(Delay_Flight))


```
```{r}
Delay_Flight$weather_delay[which(is.na(Delay_Flight$weather_delay))]<- median(Delay_Flight$weather_delay,na.rm = T)
sum(is.na(Delay_Flight))






```

```{r}
Delay_Flight$nas_ct[which(is.na(Delay_Flight$nas_ct))]<- 
median(Delay_Flight$nas_ct,na.rm = T)
sum(is.na(Delay_Flight))



```
```{r}
Delay_Flight$security_ct[which(is.na(Delay_Flight$security_ct))]<- 
median(Delay_Flight$security_ct,na.rm = T)
sum(is.na(Delay_Flight))
```

```{r}
Delay_Flight$late_aircraft_ct[which(is.na(Delay_Flight$late_aircraft_ct))]<- 
median(Delay_Flight$late_aircraft_ct,na.rm = T)
sum(is.na(Delay_Flight))
```


```{r}
Delay_Flight$arr_cancelled[which(is.na(Delay_Flight$arr_cancelled))]<- 
median(Delay_Flight$arr_cancelled,na.rm = T)
sum(is.na(Delay_Flight))
```
```{r}
Delay_Flight$nas_delay[which(is.na(Delay_Flight$nas_delay))]<- 
median(Delay_Flight$nas_delay,na.rm = T)
sum(is.na(Delay_Flight))
```

```{r}
Delay_Flight$nas_ct[which(is.na(Delay_Flight$nas_ct))]<- 
median(Delay_Flight$nas_ct,na.rm = T)
sum(is.na(Delay_Flight))
```


VISUALIZATION 

Univariate and Bivariate analysis
```{r}
#Function to plot histogram and boxplot simultaneously
plot_histogram_n_boxplot = function(variable, variableNameString, binw){
  h = ggplot(data = Delay_Flight, aes(x= variable))+
    labs(x = variableNameString,y ='count')+
    geom_histogram(fill = 'green',col = 'white',binwidth = binw)+
    geom_vline(aes(xintercept=mean(variable)),
               color="black", linetype="dashed", size=0.5)
  b = ggplot(data = Delay_Flight, aes('',variable))+ 
    geom_boxplot(outlier.colour = 'red',col = 'red',outlier.shape = 19)+
    labs(x = '',y = variableNameString)+ coord_flip()
  grid.arrange(h,b,ncol = 2)
}


```

```{r echo=FALSE}
# Dependent variable mpg

plot_histogram_n_boxplot(Delay_Flight$arr_del15,"arr_del15",2)

#The distributions of Arr_del15 is left skewed.
```



```{r echo=FALSE}
plot_histogram_n_boxplot(Delay_Flight$nas_ct,"Delay flight",300)
```



```{r echo=FALSE}
plot_histogram_n_boxplot(Delay_Flight$carrier_ct,"Carrier category",10)
```
- The distributions of carrier category is left skewed. And has outliers which can affect our linear model s we have to handle these outliers.

```{r echo=FALSE}
plot_histogram_n_boxplot(Delay_Flight$weather_delay,"Weather delay",20)
```
- The distributions for displacement is left skewed/tailed.


```{r echo=FALSE}
plot_histogram_n_boxplot(Delay_Flight$arr_cancelled,"arline cancelled",1)

```
- The distributions for arline cancelled looks left skewed distributed

```{r}
unique(Delay_Flight$year)
```


```{r}
ggplot(Delay_Flight, aes(x=carrier)) + labs(x = "Airline ",y ='Distribution') + geom_bar( fill = 'green', color ="orange") + theme_classic()
ggplot(Delay_Flight, aes(x=year)) + labs(x = "Trends of years ",y ='Distribtion') + geom_bar( fill = 'green', color ="orange") + theme_classic()


```





Multivariate Correlation plot between the numerical variables
```{r}
library(GGally)
```
```{r}
num_vars = sapply(Delay_Flight[,c(-7,-10)], is.numeric)

DataExplorer::plot_correlation(Delay_Flight[,num_vars])
```


Correlation between dependent variable and independent variable is to be considered when modelling a prediction in a dataset, but if there is high correlation amongst independent variables themselves, that causes problem in interpreting statistical models and therefore will treat them later.

Carrier_cat,weather_delay,nas_cat,security_cat has a positive correlation with the arr_del15 implying as all varable of a delay_flight increases the arr_del15  increases.

- We can see that our independent variables show a high correlation with each other contributing to multicollinearity that needs to be taken care of much later.
`

Outlier treatment

As observed earlier carrier_ct,weather_delay,nas_cat,security_cat has outlier and we need to treat them as linear models are sensitive to the effects of outliers. We will create a function to cap the outliers

```{r}
cap_outliers <- function(x, lower_cap = NULL, upper_cap = NULL) {
  # Calculate the quartiles
  q1 <- quantile(x, 0.25)
  q3 <- quantile(x, 0.75)
  
  # Calculate the interquartile range (IQR)
  iqr <- q3 - q1
  
  # Calculate the lower and upper fences
  lower_fence <- q1 - 1.5 * iqr
  upper_fence <- q3 + 1.5 * iqr
  
  # Cap outliers
  if (!is.null(lower_cap)) {
    x[x < lower_fence] <- lower_cap
  }
  
  if (!is.null(upper_cap)) {
    x[x > upper_fence] <- upper_cap
  }
  
  return(x)
}

```
Now will can apply the outlier capping function to the variable we want




```{r}
Delay_Flight$carrier_ct <- cap_outliers(Delay_Flight$carrier_ct)
Delay_Flight$weather_delay <- cap_outliers(Delay_Flight$weather_delay)
Delay_Flight$nas_ct <- cap_outliers(Delay_Flight$nas_ct)
Delay_Flight$security_delay <- cap_outliers(Delay_Flight$security_ct)
Delay_Flight$arr_flights <- cap_outliers(Delay_Flight$arr_flights)
Delay_Flight$arr_cancelled <- cap_outliers(Delay_Flight$arr_cancelled)
Delay_Flight$late_aircraft_ct <- cap_outliers(Delay_Flight$late_aircraft_ct)
```




MODEL BUILDING

# Model Building Approach :

1. Partition the data into train and test set.
2. Build a Linear Regression model on the train data.
3. Test the data on test set.


```{r}
library(caTools) # for split

# Set seed to ensure reproducibility

set.seed(12)
# use the function sample() to randomly sample 70% rows into train data and 30% to test data
sample = sample.split(Delay_Flight$arr_del15, SplitRatio = 0.7)
traindata = subset(Delay_Flight, sample == TRUE)
testdata = subset (Delay_Flight, sample == FALSE)

## View to see the distribution of the dependent variable accross the split in comparison to total distribution

```

```{r}
head(traindata)

nrow(traindata)
nrow(testdata)

```
# Creating a baseline model :
* A baseline model in Linear Regression is the average of the dependent variable as the prediction. As maybe a guide for comaprison later


Build a Linear Regression model on the train dataset

* lm is used to fit a linear model and perform regression
* formula is a symbolic description of the model to be fitted like dependent_variable~Independent_variables

```{r}
model_1 <-lm(formula =  arr_del15~. ,data=traindata)
summary(model_1)

```
-- Understanding the output of Linear Regression model 

Model performance :

- Multiple R-squared and Adjusted R-squared : It is always between 0 to 1, high value signifies better percentage of variation in the dependent variable.
When we increase number of variables the Multiple R-squared value may increase but Adj R-squared only increases when the added column is significant for explaining the variation in the dependent variable.
In our model we have got a multiple R-squared of 1 and an adjusted R-squared of 1 which looks good


-- Multi-collinearity and VIF :

- Multicollinearity occurs when predictor variables in a regression model are correlated. This correlation is a problem because predictor variables should be independent. 
If the correlation between variables is high, it can cause problems when we fit the model and interpret the results.
When we have multicollinearity the linear model, The coefficients that the model suggests are unreliable.

- To get rid of with multicollinearity we use Variance Inflation Factor (VIF)

- In order to check Multicolliniearity, we use Variance Inflation Factor(VIF) among the predictors.A value greater than 10 signifies high multicolliniearity and hence it is advisable to drop such features.


-- Checking multicollinearity in data 
```{r}

olsrr::ols_vif_tol(model_1)

```

```{r}
model_2<-lm(formula = arr_del15~ carrier_ct + weather_delay + nas_ct + security_ct    ,data=traindata)

summary(model_2)
```

```{r}
model_3<-lm(formula = arr_del15~ carrier_ct + weather_delay + nas_ct + security_ct +   +arr_cancelled+late_aircraft_ct+arr_flights,data=traindata)
summary(model_3)
```

```{r}
## create our testing input, which is the test data minus the target variable arr-del15
xtest <- testdata[,-c(1)]
```

```{r}
## create the test output, which is just the target variable from the test set
ytest <- testdata$arr_del15
```

```{r}
## Get our models prediction by using our model to predict the test inputs "xtest"
pred <- predict(model_2,xtest)
```

```{r}
## comparring the correllation between our predicted value and the actual value
## This gives us the performance of our model
cor(pred,ytest)
```

```{r}
#visualizing our model prediction alongside the actual test values
result = data.frame(pred,ytest)
cbind(pred,ytest)
```





```{r}
#Plotting the prediction against the actual test values to visualize the model performance
ggplot(result, aes(x = ytest, y = pred)) + 
  geom_point()+geom_smooth()
```
Accuracy

```{r}
RMSE <- sqrt(mean(pred-Delay_Flight$arr_del15)^2)


